Automatic Differentiation Accelerates Inference for Partially Observed Stochastic Processes

### 2024-12-07 paper/ms.tex
810 presentation

\begin{figure}[t!]
    \centering
    \vspace{1mm}
    \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth}/\real{1.25}]{imgs/pmcmc/nuts_eb.png}
    \caption{Histogram and density estimate of the posterior of $\beta_{\text{trend}}$, as well as convergence diagnostics for NUTS with $4$ chains. NUTS mixes quickly when regularized by an empirical prior with a standard deviation of $6.3\%$ of the MLE, and the posterior estimates from each chain share roughly the same posterior mode.}
    \label{fig:nuts-eb}
    \arxiv{}{\vspace*{-4mm}}
\end{figure}

>>>>

\begin{figure}[t!]
    \centering
    \vspace{1mm}
    \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth}/\real{1.25}]{imgs/pmcmc/nuts_eb.png}
    \caption{Convergence diagnostics for NUTS with $4$ chains, where we plot $1000$ samples from the posterior of $\beta_{\text{trend}}$ within the Dhaka cholera model of \cite{king08}. \textbf{A:} Histogram of posterior samples. \textbf{B:} density estimate of the posterior. \textbf{C:} Autocorrelation plot for MCMC samples. \textbf{D:} Trace plot for MCMC samples, pre and post burn-in. NUTS mixes quickly when regularized by an empirical prior with a standard deviation of $6.3\%$ of the MLE, and the posterior estimates from each chain share roughly the same posterior mode.}
    \label{fig:nuts-eb}
    \arxiv{}{\vspace*{-4mm}}
\end{figure}

### 2024-12-07 presentation/810class14.tex
810 presentation

\documentclass{beamer}
\setbeamertemplate{caption}[numbered]
\usepackage{tikz}
\usepackage{graphicx}
\usetheme{AnnArbor}
%\usecolortheme{wolverine}
\title[AD for POMPs]{Automatic Differentiation Accelerates Inference for Partially Observed Stochastic Processes}
\subtitle{\small{Inference for Partially Observed Structured Dynamic Systems}}
\author[Kevin Tan]{Kevin Tan \\ \small{University of Pennsylvania, Department of Statistics and Data Science} \\\vspace{0.5cm} \small{Joint work with Giles Hooker and Edward Ionides}}
\date[January 4, 2024]{IMS Asia Pacific Rim Meeting \\ January 4, 2024}

\usepackage{amsmath}
\usepackage{tabularx}
\usepackage{mathtools}
\usepackage{bbold}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{pifont}
%\usepackage[left=1in,right=1in,]{geometry}
\usepackage{multicol,lipsum,xcolor}
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{amsmath,amsthm,amssymb,bm}
\usepackage{bbold}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{enumerate}
%\usepackage{enumitem}
\usepackage{natbib}
\usepackage{parskip}
\usepackage{comment}
\usepackage{microtype}
\usepackage{diagbox}
\usepackage{tikz}
\usepackage{fancyhdr}
\usepackage{appendixnumberbeamer}

@@

\include{mop/macros}

@@

    \begin{figure}
        \centering
        \begin{minipage}[c]{0.5\textwidth}
            \centering
            \includegraphics[scale=0.15]{imgs/pfilter.png}
            \end{minipage}
        \begin{minipage}[c]{0.45\textwidth}
            \caption{
            \\
            -- Dirac measure approximation of the Bayes filter. 
            \\ 
            -- Yields a filtering distribution and likelihood estimate.
            \\ 
            -- Graphic from \cite{berg2019pfilterimage}.}
        \end{minipage}
        \label{fig:pfilter-illustration}
    \end{figure}

@@

    \begin{figure}
        \begin{minipage}[c]{0.7\textwidth}
            \includegraphics[scale=0.3]{imgs/autograd.png}
        \end{minipage}
        \begin{minipage}[l]{0.27\textwidth}
            \caption{
            \\$f(x,y) = x^2+xy+2$, $f'(x,y) = (2x+y,x).$
            -- Figure from \href{https://avinashselvam.medium.com/automatic-differentiation-explained-9f02c74e9a90}{Medium}.}
        \end{minipage}
        \label{fig:autograd-illustration}
    \end{figure}

@@

\begin{enumerate}
    \item \textbf{Intractable likelihood functions}! Bypassed with e.g. simulated likelihood, likelihood-free inference, particle filters, etc. 
    \item \pause May not have access to transition densities, only a \textbf{simulator}. 
    \begin{itemize}
        \item \pause IF2 (\cite{ionides15}), particle MCMC (\cite{doucet2010pmcmc}), only full-information methods that can deal with this.
    \end{itemize}
    \item \pause \textbf{Significant Monte-Carlo noise in likelihood estimate} makes accurate parameter estimation difficult.
\end{enumerate}
\end{frame}

@@

\begin{frame}{Hang on, wait a minute!}
    \begin{itemize}
        \item Original iterated filtering algorithm: (noisy) approximation of score (\cite{ionides06-pnas}). 
        %\item \pause Performed a single "gradient step" at the end of each filtering iteration.
        \pause
        \item \textbf{Would performing automatic differentiation (AD) on the likelihood estimate from the particle filter lead to a less noisy score approximation?}
    \end{itemize}
    \pause
    \begin{block}{Problem!}
        But the particle filter has discrete stochastic resampling! How can we differentiate this?
    \end{block}
\end{frame}

@@

    \begin{table}[h!]
        \centering
        \begin{tabular}{||c c||} 
         \hline
         Authors & Method \\ [0.5ex] 
         \hline\hline
         \cite{poyiadjis11} & Particle approximation of score
         \\ & $O(N^2)$ variance, unbiased \\
         \hline
         \cite{blei2018vsmc} & Backprop through vanilla PF \\ 
         & Asymptotic bias \\
         \hline
         \cite{corenflos21} & Optimal transport resampling \\
         & Consistent, $O(J^2)$ runtime \\
         \hline
         \cite{scibior21} & Stop-gradient trick \\ 
         & Recovers \cite{poyiadjis11} with AD \\
         \hline 
         \cite{singh22} & Fixed-lag smoothing \\ 
         & Need transition densities \\ 
         \hline
        \end{tabular}
        \label{table:lit-review}
    \end{table}
\end{frame}

\begin{frame}{Our Contributions}
    \begin{itemize}
        \item New theoretical framework/algorithm/gradient estimator we call MOP-$\alpha$.
        \begin{itemize}
            \item \pause Gradient estimates of \cite{blei2018vsmc} (MOP-0), \cite{poyiadjis11} and \cite{scibior2021dpf} (MOP-1) are special cases.
            \item \pause Does not need transition densities, only a differentiable simulator.
            \item \pause Can optimize a bias-variance tradeoff. 
        \end{itemize}
        \item \pause Promising hybrid algorithm, warm-starts gradient descent (using this estimator) with IF2.
        \item \pause Outperforms IF2 on Cholera model of \cite{king08}.
    \end{itemize}
\end{frame}

@@

\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
\begin{itemize}
    \item $\alpha$ balances a tradeoff between:
    \begin{itemize}
        \item \pause Maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$).
        \item \pause Considering only the single-step transition dynamics (when $\alpha=0$).
    \end{itemize}
    \item \pause\textbf{Exponentially-weighted moving average} where $\alpha$ controls the amount of discounting.
    \item \pause Bias small if $y_{n+1:N}^*$ not informative for $x_n$ given $y_{0:n}^*$ (\cite{corenflos21}).
\end{itemize}

\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
\begin{figure}
    \centering
    \includegraphics[scale=0.5]{imgs/mlegrad.png}
    \caption{Gradient estimates at MLE for trend in \cite{king08} cholera model. When $\alpha=1$, estimate close to unbiased with high variance. When $\alpha=0$, lower variance but biased.}
    \label{fig:bias-variance}
\end{figure}

@@

\begin{frame}{Guarantee}
    \begin{prop}[Correctness of MOP-$\alpha$]
    \label{prop:mop-correctness}
    When either $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the posterior and is unbiased for the likelihood under $\theta$. 
    When $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter. Its gradient when $\alpha=1$ is the estimate of \citet{poyiadjis11},
    $$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^*|x_{1:N,j}^{A, F,\theta}),$$
    and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
    $$
        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
    $$
    i.e. differentiating through a vanilla particle filter (\cite{scibior21}).
\end{prop}
\end{frame}

@@

\begin{frame}{Convergence Analysis}
    \begin{itemize}
        \item Similar to \cite{mahoney16}, uses concentration inequalities from \cite{delmoral2011ci}.
        \item Gradient stage of IFAD-1 converges linearly to the MLE if:
        \begin{enumerate}
            \item \pause The log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE.
            \item \pause The IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE. 
        \end{enumerate}
        \item \pause \textbf{Conjecture:} This applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE and behaves a lot like SGD. 
    \end{itemize}

@@

\begin{itemize}
    \item Warm-start:
    \begin{itemize}
        \item \pause Initial warm-start "convergence" happens fairly quickly in practice. 
        \begin{itemize}
            \item \pause Dhaka cholera model of \cite{king08}: 40 iterations, usually 100-200 for global search.
        \end{itemize}
    \end{itemize}
    \item \pause MOP-$\alpha$:
    \begin{itemize}
        \item \pause Gradient takes 3.75x time of \texttt{pfilter()}, in line with cheap gradient principle (\cite{kakade2019provably}).
    \end{itemize}
    \item \pause Runtime \textbf{linear, and not quadratic}, in number of particles.
    \item \pause Re-implementation in \texttt{JAX} (\cite{jax2018github}) led to 16x speedup v.s. \texttt{pomp} package of \cite{king08}.
\end{itemize}

@@

\begin{frame}{Future Work}
    \begin{itemize}
        \item MOP-$\alpha$ cannot handle discrete latent states. Maybe likelihood ratios can get around this.
        \item \pause Convergence rate of IF2 not known, conjecture similar behavior to SGD. 
        \item \pause Only gradient descent explored, other methods like Newton's method, ADAM, L-BFGS, etc. may perform better.
        \item \pause Extensions to panel and spatiotemporal data, e.g. with a block particle filter in the lens of \cite{ionides22} and \cite{ning23}.
        \item \pause Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual}.
    \end{itemize}
\end{frame}

\bibliographystyle{apalike}
\bibliography{mop/bib-ifad, mop/bib-ref}

@@

    \begin{itemize}
        \item \cite{poyiadjis11}, particle approximation of the score. \pause \\ \cite{scibior21}: AD + "stop-gradient trick". 
        \item 
        \pause \cite{blei2018vsmc}, backpropagate through vanilla particle filter.
        \item 
        \pause \cite{corenflos21}, optimal transport resampling.
        \item 
        \pause \cite{singh22}, fixed-lag smoothing.
    \end{itemize}
\end{frame}

\begin{frame}{Previous Work: Issues}
    \begin{itemize}
        \item Not (yet) compatible with desire for simulation-based inference.
        \begin{itemize}
            \item \pause Need modifications to use the bootstrap filter.
            \item \pause Sometimes special cases needed, e.g. transitions factoring into a policy and deterministic model (\cite{singh22}).
        \end{itemize}
        \item \pause Computationally expensive, quadratic in number of particles.
        \begin{itemize}
            \item \pause Optimal transport resampling (\cite{corenflos21}), marginal particle filters (\cite{scibior21}).
        \end{itemize}
        \item \pause High variance, or asymptotically biased.
        \begin{itemize}
            \item \pause Either one drops resampling terms and accepts asymptotic bias (\cite{blei2018vsmc}), or has variance quadratic in horizon (\cite{poyiadjis11}, \cite{scibior21})
        \end{itemize}
    \end{itemize}
\end{frame}

>>>>

\documentclass{beamer}
\setbeamertemplate{caption}[numbered]
\usepackage{tikz}
\usepackage{graphicx}
\usetheme{AnnArbor}
%\usecolortheme{wolverine}
\title[AD for POMPs]{Automatic Differentiation Accelerates Inference for Partially Observed Stochastic Processes}
\subtitle{\url{https://arxiv.org/abs/2407.03085}}
\author[Tan, Hooker \& Ionides]{Kevin Tan$^1$, Giles Hooker$^1$ and Ed Ionides$^2$ \\ {\small $^1$University of Pennsylvania, $^2$University of Michigan}} 
\date[STATS 810]{STATS 810 \\December 9, 2024}

\usepackage{amsmath}
\usepackage{tabularx}
\usepackage{mathtools}
\usepackage{bbold}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{pifont}
%\usepackage[left=1in,right=1in,]{geometry}
\usepackage{multicol,lipsum,xcolor}
%\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{amsmath,amsthm,amssymb,bm}
\usepackage{bbold}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{enumerate}
%\usepackage{enumitem}
\usepackage{natbib}
\usepackage{parskip}
\usepackage{comment}
\usepackage{microtype}
\usepackage{diagbox}
\usepackage{tikz}
\usepackage{fancyhdr}
\usepackage{appendixnumberbeamer}

@@

\input{mop/macros}

@@

    \begin{figure}
        \centering
        \begin{minipage}[c]{0.5\textwidth}
            \centering
            \includegraphics[scale=0.15]{imgs/pfilter.png}
            \end{minipage}
        \begin{minipage}[c]{0.45\textwidth}
            \caption{
            \\
            -- Dirac measure approximation of the Bayes filter. 
            \\ 
            -- Yields a filtering distribution and likelihood estimate.
            \\ 
            -- Graphic from \cite{berg19}.}
        \end{minipage}
        \label{fig:pfilter-illustration}
    \end{figure}

@@

    \begin{figure}
        \begin{minipage}[c]{0.7\textwidth}
            \includegraphics[scale=0.3]{imgs/autograd.png}
        \end{minipage}
        \begin{minipage}[l]{0.27\textwidth}
            \caption{
            \\$f(x,y) = x^2+xy+2$, $f'(x,y) = (2x+y,x).$
            -- Figure from             %\url{https://avinashselvam.medium.com/automatic%-differentiation-explained-9f02c74e9a90}  
            \url{https://avinashselvam.medium.com}
            }
        \end{minipage}
        \label{fig:autograd-illustration}
    \end{figure}

@@

\begin{enumerate}
    \item \textbf{Intractable likelihood functions}! Bypassed with e.g. simulated likelihood, likelihood-free inference, particle filters, etc. 
    \item \pause May not have access to transition densities, only a \textbf{simulator}. 
    \begin{itemize}
        \item \pause IF2 \citep{ionides15}, particle MCMC \citep{andrieu10}, only full-information methods that can deal with this.
    \end{itemize}
    \item \pause \textbf{Significant Monte-Carlo noise in likelihood estimate} makes accurate parameter estimation difficult.
\end{enumerate}
\end{frame}

@@

\begin{frame}{Hang on, wait a minute!}
    \begin{itemize}
        \item Original iterated filtering algorithm: (noisy) approximation of score \citep{ionides06-pnas}. 
        %\item \pause Performed a single "gradient step" at the end of each filtering iteration.
        \pause
        \item \textbf{Would performing automatic differentiation (AD) on the likelihood estimate from the particle filter lead to a less noisy score approximation?}
    \end{itemize}
    \pause
    \begin{block}{Problem!}
        But the particle filter has discrete stochastic resampling! How can we differentiate this?
    \end{block}
\end{frame}

@@

    \begin{table}[h!]
        \centering
        \begin{tabular}{||c c||} 
         \hline
         Authors & Method \\ [0.5ex] 
         \hline\hline
         \cite{poyiadjis11} & Particle approximation of score
         \\ & $O(N^2)$ variance, unbiased \\
         \hline
         \cite{naesseth18} & Backprop through vanilla PF \\ 
         & Asymptotic bias \\
         \hline
         \cite{corenflos21} & Optimal transport resampling \\
         & Consistent, $O(J^2)$ runtime \\
         \hline
         \cite{scibior21} & Stop-gradient trick \\ 
         & Recovers \cite{poyiadjis11} with AD \\
         \hline 
         \cite{singh22} & Fixed-lag smoothing \\ 
         & Need transition densities \\ 
         \hline
        \end{tabular}
        \label{table:lit-review}
    \end{table}
\end{frame}

\begin{frame}{Our Contributions}
    \begin{itemize}
        \item New theoretical framework/algorithm/gradient estimator we call MOP-$\alpha$.
        \begin{itemize}
            \item \pause Gradient estimates of \cite{naesseth18} (MOP-0), \cite{poyiadjis11} and \cite{scibior21} (MOP-1) are special cases.
            \item \pause Does not need transition densities, only a differentiable simulator.
            \item \pause Can optimize a bias-variance tradeoff. 
        \end{itemize}
        \item \pause Promising hybrid algorithm, warm-starts gradient descent (using this estimator) with IF2.
        \item \pause Outperforms IF2 on Cholera model of \cite{king08}.
    \end{itemize}
\end{frame}

@@

\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
\begin{itemize}
    \item $\alpha$ balances a tradeoff between:
    \begin{itemize}
        \item \pause Maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$).
        \item \pause Considering only the single-step transition dynamics (when $\alpha=0$).
    \end{itemize}
    \item \pause\textbf{Exponentially-weighted moving average} where $\alpha$ controls the amount of discounting.
    \item \pause Bias small if $y_{n+1:N}^*$ not informative for $x_n$ given $y_{0:n}^*$ \citep{corenflos21}.
\end{itemize}

@@

\begin{frame}{Guarantee}
    \begin{prop}[Correctness of MOP-$\alpha$]
    \label{prop:mop-correctness}
    When either $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the posterior and is unbiased for the likelihood under $\theta$. 
    When $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter. Its gradient when $\alpha=1$ is the estimate of \citet{poyiadjis11},
    $$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^*|x_{1:N,j}^{A, F,\theta}),$$
    and when $\alpha=0,$ is the gradient estimator of \cite{naesseth18},
    $$
        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
    $$
    i.e. differentiating through a vanilla particle filter \citep{scibior21}.
\end{prop}
\end{frame}

@@

\begin{frame}{Convergence Analysis}
    \begin{itemize}
        \item Similar to \cite{mahoney16}, uses concentration inequalities from \cite{delmoral11}.
        \item Gradient stage of IFAD-1 converges linearly to the MLE if:
        \begin{enumerate}
            \item \pause The log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE.
            \item \pause The IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE. 
        \end{enumerate}
        \item \pause \textbf{Conjecture:} This applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE and behaves a lot like SGD. 
    \end{itemize}

@@

\begin{itemize}
    \item Warm-start:
    \begin{itemize}
        \item \pause Initial warm-start "convergence" happens fairly quickly in practice. 
        \begin{itemize}
            \item \pause Dhaka cholera model of \cite{king08}: 40 iterations, usually 100-200 for global search.
        \end{itemize}
    \end{itemize}
    \item \pause MOP-$\alpha$:
    \begin{itemize}
        \item \pause Gradient takes 3.75x time of \texttt{pfilter()}, in line with cheap gradient principle \citep{kakade2019provably}.
    \end{itemize}
    \item \pause Runtime \textbf{linear, and not quadratic}, in number of particles.
    \item \pause Re-implementation in \texttt{JAX} \citep{jax} led to 16x speedup v.s. \texttt{pomp} package of \cite{king08}.
\end{itemize}

@@

\begin{frame}{Future Work}
    \begin{itemize}
        \item MOP-$\alpha$ cannot handle discrete latent states. Maybe likelihood ratios can get around this.
        \item \pause Convergence rate of IF2 not known, conjecture similar behavior to SGD. 
        \item \pause Only gradient descent explored, other methods like Newton's method, ADAM, L-BFGS, etc. may perform better.
        \item \pause Extensions to panel and spatiotemporal data, e.g. with a block particle filter in the lens of \cite{ionides22} and \cite{ning23}.
        \item \pause Python counterpart to the popular \texttt{pomp} R package by \citet{king16}.
    \end{itemize}
\end{frame}

\bibliographystyle{apalike}
\bibliography{paper/bib-ifad}

@@

    \begin{itemize}
        \item \cite{poyiadjis11}, particle approximation of the score. \pause \\ \cite{scibior21}: AD + "stop-gradient trick". 
        \item 
        \pause \cite{naesseth18}, backpropagate through vanilla particle filter.
        \item 
        \pause \cite{corenflos21}, optimal transport resampling.
        \item 
        \pause \cite{singh22}, fixed-lag smoothing.
    \end{itemize}
\end{frame}

\begin{frame}{Previous Work: Issues}
    \begin{itemize}
        \item Not (yet) compatible with desire for simulation-based inference.
        \begin{itemize}
            \item \pause Need modifications to use the bootstrap filter.
            \item \pause Sometimes special cases needed, e.g. transitions factoring into a policy and deterministic model (\cite{singh22}).
        \end{itemize}
        \item \pause Computationally expensive, quadratic in number of particles.
        \begin{itemize}
            \item \pause Optimal transport resampling (\cite{corenflos21}), marginal particle filters (\cite{scibior21}).
        \end{itemize}
        \item \pause High variance, or asymptotically biased.
        \begin{itemize}
            \item \pause Either one drops resampling terms and accepts asymptotic bias (\cite{naesseth18}), or has variance quadratic in horizon (\cite{poyiadjis11}, \cite{scibior21})
        \end{itemize}
    \end{itemize}
\end{frame}
